---
title: "Power usage analysis"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 1
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```
Latest update: `r format(Sys.Date(), "%d.%m.%Y")`

Welcome to my power usage project! Through the fantastic service [elhub](https://www.elhub.no/) I have downloaded (and will continue to download) monthly data sets containing the hourly power usage in my appartment. I made an [R-package](https://github.com/emmaSkarstein/power.usage.analysis) that contains all the data plus some useful functions. Here I will explore some visualisations of the data.


```{r, warning=FALSE, message=FALSE}
library(power.usage.analysis)
library(ggplot2)
library(dplyr)

theme_set(theme_minimal())
```


First, let us load the data and look at the structure of it.


```{r, warning=FALSE, message=FALSE}
data <- clean_and_prepare(power_data)
data$hour <- format(data$tid, format = "%H:%M:%S")
head(data)
```

# Basic plot

Let's try a simple plot of all the observations.
```{r Messy plot, warning=FALSE, message=FALSE}
p_simple <- ggplot(data, aes(x = tid, y = forbruk)) + 
  geom_line() +
  ylab("Power usage") +
  xlab("")
p_simple
```

This is a mess. While we do see some general tendencies that are interesting (like maximum consumption varying seasonally), it is difficult to tell what is going on in all the noise. Maybe we can smooth it?

```{r Smoothed plot, warning=FALSE, message=FALSE}
p_simple + geom_smooth(color = "darkorange")
```

This certainly looks smooth, but it's still looking quite messy. I'll try to add a better theme than the default.

----add theme----


-----also comment on missing values------


# Average per weekday

Another option is to try to group the data by some time period by summing or averaging it.

Let's first see if there seems to be any difference in consumption for each weekday. 
```{r Average per weekday, warning=FALSE, message=FALSE}
ggplot(data, aes(x = weekday, y = forbruk)) +
  geom_violin(fill = "darkorange") +
  geom_boxplot(width=0.1) +
  ylab("Power usage") +
  xlab("")
```

Okay, it pretty much looks like the distribution is the same for each week day, maybe not surprisingly, although I might have expected the consumption to be a bit higher during the week-end. However, during most of 2020 we worked from home. Could it be that this is affecting the results, and that weekday differences were larger in 2019? Let's investigate:

```{r Average per weekday by year, warning=FALSE, message=FALSE}
ggplot(data, aes(x = weekday, y = forbruk, fill = year)) +
  geom_violin() +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Power usage") +
  xlab("")

ggplot(data, aes(x = weekday, y = forbruk, fill = year)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Power usage") +
  xlab("")
```


# Total per month

```{r, warning=FALSE, message=FALSE}
monthly_data <- data %>%
  group_by(month) %>%
  summarise(sum_forb = sum(forbruk))

year_month <- data %>%
  group_by(month, year) %>%
  summarise(sum_forb = sum(forbruk))
```


```{r Total per month, warning=FALSE, message=FALSE}
ggplot(year_month, aes(x = month, y = sum_forb)) +
  geom_col(aes(fill = year), color = "black", position = "dodge") +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Total power usage per month") +
  xlab("")
```


# Usage throughout the day

```{r}
ggplot(data, aes(x = hour, y = forbruk, fill = year)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Power usage") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45))
```


